---
title: "Understanding Probationary Periods in De Corte's Utility Analysis"
subtitle: "A Detailed Explanation of the Selection-Retention Framework"
author: "Reproduction Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)

# Load the results for demonstration
results <- read.csv("de_corte_1994_results.csv")
```

# Introduction

De Corte's (1994) utility analysis framework introduces a crucial innovation: **probationary periods** in selection decisions. This document explains how the model accounts for the reality that not all selected employees will be retained, and how this affects utility calculations.

## The Problem with Traditional Utility Analysis

Traditional utility analysis (e.g., Brogden, 1949) assumes that:
- All selected employees remain on the job
- Performance is immediately observable
- No costs are associated with employee separation

However, in practice:
- **Probationary periods** are common (3-6 months)
- **Some employees fail probation** and are dismissed
- **Separation costs** are incurred for unsuccessful employees
- **Performance is only observed after probation**

# The Probationary Period Framework

## Key Concepts

### 1. Two-Stage Decision Process

The model conceptualizes selection as a **two-stage process**:

1. **Selection Stage**: Choose N employees from n applicants using a predictor
2. **Retention Stage**: After probation, retain only employees who meet performance standards

### 2. Critical Values

Two critical values determine the process:

- **x_c**: Critical predictor score (determines who gets hired)
- **r_c**: Critical performance score (determines who gets retained)

### 3. Success Ratios

- **S_p**: Success ratio of the predictor (proportion of selectees who pass probation)
- **S_0**: Success ratio of random selection (proportion who would pass probation if selected randomly)

## Mathematical Framework

### Basic Utility Formula

The utility of predictor-based selection with probationary periods is:

$$U_p = N \times (\mu_y(x_c) - \mu_s) + (T-1) \times N \times S_p \times (\mu_y(x_c,r_c) - \mu_s) - N \times C_t - n \times C_p$$

Where:
- $N$ = Number of hires
- $\mu_y(x_c)$ = Average payoff of all selectees
- $\mu_s$ = Average service cost
- $T$ = Average time periods on job for successful employees
- $S_p$ = Success ratio (proportion who pass probation)
- $\mu_y(x_c,r_c)$ = Average payoff of successful selectees
- $C_t$ = Training cost per hire
- $C_p$ = Cost per candidate for predictor

### Key Components Explained

#### 1. Initial Period Utility
$$N \times (\mu_y(x_c) - \mu_s)$$

This represents the utility from all selected employees during the probationary period (Period 1).

#### 2. Retention Period Utility
$$(T-1) \times N \times S_p \times (\mu_y(x_c,r_c) - \mu_s)$$

This represents the utility from retained employees over the remaining periods (Periods 2 to T).

#### 3. Costs
$$-N \times C_t - n \times C_p$$

Training costs for all hires plus predictor costs for all applicants.

# How Probationary Periods Affect Utility

## 1. Success Ratio Impact

The success ratio $S_p$ is crucial because it determines:
- How many employees actually contribute to long-term utility
- The effective workforce size after probation
- The economic impact of selection accuracy

```{r success-ratio-impact, echo=FALSE}
# Create example to show success ratio impact
success_example <- data.frame(
  Scenario = c("Random Selection", "Predictor Selection"),
  Success_Ratio = c(0.663, 0.853),
  Employees_Hired = c(17, 17),
  Employees_Retained = c(17 * 0.663, 17 * 0.853),
  Retained_Count = c(11.27, 14.50)
)

kable(success_example, 
      col.names = c("Scenario", "Success Ratio", "Employees Hired", "Employees Retained", "Retained Count"),
      caption = "Impact of Success Ratio on Employee Retention") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 2. Conditional Performance Expectations

The model uses **conditional expectations** to account for the correlation between selection and retention:

### μ_y(x_c) - Average Performance of All Selectees
This represents the expected performance of all employees during probation.

### μ_y(x_c,r_c) - Average Performance of Successful Selectees
This represents the expected performance of employees who pass probation.

### μ_y(r_c) - Average Performance of Random Successful Employees
This represents the expected performance if selection were random.

## 3. Time Period Structure

The model distinguishes between:
- **Period 1**: Probationary period (all selectees)
- **Periods 2-T**: Post-probation periods (only retained employees)

This creates a **two-tier utility structure**:
1. Initial investment period (probation)
2. Return period (retention)

# Practical Implications

## 1. Economic Realism

The model accounts for real organizational costs:
- **Training costs** for all hires
- **Separation costs** for unsuccessful employees
- **Recruitment costs** for different selection methods

## 2. Decision Making Scenarios

### Fixed Quota of Selectees
When the goal is to hire a fixed number of employees:
- Calculate how many need to be hired to achieve target retention
- Compare costs of different selection methods

### Fixed Quota of Successful Employees
When the goal is to have a fixed number of successful employees:
- Calculate required hiring numbers for each method
- Account for separation and recruitment costs

## 3. Sensitivity to Key Parameters

The model is particularly sensitive to:
- **Predictor validity** (ρ)
- **Success ratio** (S_p)
- **Performance variability** (σ_y)
- **Time horizon** (T)

# Comparison with Traditional Models

## Traditional Brogden Model
$$U = N \times T \times (Z_{barx} \times r \times SD_y) - C$$

**Limitations:**
- No probationary period
- No separation costs
- No retention uncertainty

## De Corte's Enhanced Model
**Advantages:**
- Realistic probationary periods
- Separation cost accounting
- Retention probability modeling
- Two-stage decision framework

# Example from De Corte (1994)

Using the parameters from the original paper:

```{r example-calculation, echo=FALSE}
# Display key results from our reproduction
example_results <- data.frame(
  Metric = c("Employees Hired (N)", "Success Ratio (S_p)", "Employees Retained", 
             "Utility with Predictor", "Utility with Random Selection", "Utility Difference"),
  Value = c(17, 0.853, round(17 * 0.853, 2), 
            paste0("$", format(round(results$Value[6]), big.mark=",")),
            paste0("$", format(round(results$Value[7]), big.mark=",")),
            paste0("$", format(round(results$Value[8]), big.mark=",")))
)

kable(example_results, 
      col.names = c("Metric", "Value"),
      caption = "Example Results from De Corte (1994)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Interpretation

1. **17 employees are hired** using the predictor
2. **14.50 employees (85.3%) pass probation** and are retained
3. **Utility difference of $459,566** compared to random selection
4. **Economic value** comes from both better initial selection and higher retention rates

# Key Insights

## 1. Probationary Periods Reduce Utility
- Not all selected employees contribute to long-term value
- Training costs are incurred for employees who don't stay
- Success ratios significantly impact overall utility

## 2. Predictor Validity Matters More
- Higher validity improves both selection and retention
- The correlation between predictor and performance affects retention rates
- Small improvements in validity can have large utility impacts

## 3. Time Horizon is Critical
- Longer time horizons increase the value of retention
- The (T-1) multiplier amplifies retention effects
- Organizations with longer employee tenures benefit more from good selection

## 4. Cost Structure Matters
- Separation costs reduce utility differences
- Training costs affect both selection methods equally
- Recruitment costs vary by selection method

# Practical Applications

## 1. Organizational Decision Making
- **When to use probationary periods**: High training costs, uncertain performance
- **How long probation should be**: Balance between information gain and costs
- **Selection method choice**: Consider both selection and retention effects

## 2. Utility Analysis Best Practices
- **Include probationary periods** in utility calculations
- **Account for separation costs** realistically
- **Model retention probabilities** based on predictor validity
- **Consider time horizons** in utility estimates

## 3. Research Implications
- **Test predictor-retention relationships** empirically
- **Validate success ratio estimates** in different contexts
- **Explore modern selection methods** with probationary frameworks

# Conclusion

De Corte's (1994) model provides a **realistic framework** for utility analysis that accounts for the common organizational practice of probationary periods. By incorporating:

- **Two-stage decision processes**
- **Success ratios and retention probabilities**
- **Conditional performance expectations**
- **Realistic cost structures**

The model offers a more accurate and practical approach to utility analysis that reflects real organizational decision-making contexts.

The key insight is that **probationary periods fundamentally change the economics of selection** by introducing uncertainty about retention and requiring consideration of both selection and retention effects in utility calculations.

---

**Key Takeaway**: Probationary periods don't just add complexity—they fundamentally change the utility calculus by introducing retention uncertainty and requiring consideration of both selection accuracy and retention probability in decision-making. 