---
title: "Extending Utility Analysis: The Evolution of the Latham & Whyte Case"
author: "Utility Analysis Research Project"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(ggplot2)
library(knitr)
library(kableExtra)
load("latham_whyte_comprehensive_results.RData")
```

# Executive Summary

This report examines the classic Latham & Whyte (1994) budget analyst selection case using four methodological approaches, demonstrating how recent advances in utility analysis affect interpretation of classic findings.

## Four Approaches Compared

1. **Traditional utility analysis** [@brogden1949when; @cronbach1965psychological]
2. **Sturman (2000) Monte Carlo adjustments** [@sturman2000implications] 
3. **Star performer approach** [@joo2022hrm; @sturman2023getting]
4. **Combined approach** (integrating both insights)

## Key Results

- **Traditional:** $`r formatC(round(traditional$total_utility), format = "d", big.mark = ",")`
- **Sturman adjusted:** $`r formatC(round(sturman$total_utility), format = "d", big.mark = ",")` (35% reduction)
- **Star performer:** $`r formatC(round(star_power$total_utility), format = "d", big.mark = ",")` (175% increase)
- **Combined:** $`r formatC(round(combined$total_utility), format = "d", big.mark = ",")` (140% increase)

# Background: The Latham & Whyte (1994) Case

@latham1994futility examined manager reactions to utility analysis by presenting the financial results of implementing a cognitive ability test for budget analyst selection. Despite showing a $59.7 million return (14,000% ROI), managers were actually *less* likely to implement the selection procedure, suggesting the estimates lacked credibility.

## Case Study Parameters

The case involved:

- **Position:** Budget Analyst (Government agency)
- **Candidates selected:** `r format(lw_params$n_selected, big.mark = ",")`
- **Total applicants:** `r format(lw_params$n_applicants, big.mark = ",")`
- **Selection ratio:** `r paste0(lw_params$selection_ratio * 100, "%")`
- **Test validity:** `r lw_params$validity`
- **Mean salary:** $`r formatC(lw_params$mean_salary, format = "d", big.mark = ",")`
- **Time horizon:** `r lw_params$time_horizon` years

# Methodological Evolution

## Traditional Utility Analysis

The classical approach uses the Brogden-Cronbach-Gleser formula [@brogden1949when; @cronbach1965psychological]:

$$U = N_s \times SDy \times r \times \frac{\phi}{p} \times T - N_a \times C_a$$

Where SDy = 0.40 × mean salary, assuming normal performance distributions.

## Sturman (2000) Monte Carlo Insights

@sturman2000implications conducted comprehensive Monte Carlo simulations revealing that traditional utility analysis often produces unrealistically high estimates. His analysis of the original Latham & Whyte case showed:

**Sturman's Actual Findings:**
- **Original estimate:** $59,657,532
- **After all five adjustments:** Mean = $2,228,170 (**96.3% reduction**)
- **Median effect size of total adjustments:** 291%
- **Key adjustments:** Economic variables, multiple devices, top-down hiring, probationary period, employee flows

**Our Simplified Adjustment:**
- We applied a conservative **35% reduction** factor
- This represents a simplified approximation of Sturman's complex, scenario-dependent adjustments
- Sturman's actual reductions were much more dramatic (often >90%) but highly variable

## Star Performer Recognition

@joo2022hrm developed methodology recognizing that job performance follows heavy-tailed distributions where star performers create disproportionate value. @sturman2023getting confirmed their approach uses:

$$SDy_{star} = 1.1 \times \text{mean salary}$$

Based on @burke1986comparison empirical finding that SDy/SDO ratio averages 2.75.

# Comparative Analysis

```{r methodology-table}
# Table 1a: Methodology Comparison
methodology_table <- data.frame(
  Approach = c("Traditional", "Sturman", "Star Power", "Combined"),
  SDy_Method = c("40% rule", "40% rule (adjusted)", "110% rule", "110% rule (conservative)"),
  SDy_Value = summary_df$SDy_Value,
  Key_Assumption = c("Normal performance", "Range restriction", "Heavy-tailed distributions", "Balanced realism")
)

kable(methodology_table,
      col.names = c("Approach", "SDy Method", "SDy Value", "Key Assumption"),
      caption = "Table 1a: Methodological Approaches Comparison",
      align = c("l", "l", "r", "l")) %>%
  kable_styling(font_size = 10, latex_options = c("striped", "hold_position"))
```

```{r results-table}
# Table 1b: Financial Results Comparison
results_table <- data.frame(
  Approach = c("Traditional", "Sturman", "Star Power", "Combined"),
  Total_Utility = summary_df$Total_Utility,
  Per_Hire = summary_df$Per_Hire,
  Change = summary_df$Change
)

kable(results_table,
      col.names = c("Approach", "Total Utility", "Per-Hire Value", "Change (%)"),
      caption = "Table 1b: Financial Results Comparison",
      align = c("l", "r", "r", "r"),
      format.args = list(big.mark = ",")) %>%
  kable_styling(font_size = 10, latex_options = c("striped", "hold_position"))
```

```{r utility-plot, fig.cap="Total Utility Estimates by Methodological Approach", fig.height=5, fig.width=8}
plot_data <- data.frame(
  Approach = c("Traditional", "Sturman", "Star Power", "Combined"),
  Utility = c(traditional$total_utility, sturman$total_utility, 
              star_power$total_utility, combined$total_utility)
)

ggplot(plot_data, aes(x = reorder(Approach, Utility), y = Utility)) +
  geom_col(fill = "steelblue", alpha = 1.0, color = "black", size = 0.5) +
  geom_text(aes(label = paste0("$", round(Utility/1e6, 1), "M")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-6, suffix = "M"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Total Utility Estimates by Approach",
       x = "Methodological Approach", y = "Total Utility (Millions)") +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "grey90", size = 0.5),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.8),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    panel.border = element_rect(color = "black", fill = NA, size = 0.8)
  )
```

# Detailed Analysis

## Traditional Approach
- **Total Utility:** $`r formatC(round(traditional$total_utility), format = "d", big.mark = ",")`
- **Per-Hire Value:** $`r formatC(round(traditional$per_hire_utility), format = "d", big.mark = ",")`
- **SDy:** $`r formatC(round(traditional$sdy), format = "d", big.mark = ",")` (40% rule)

Uses the classical Brogden-Cronbach-Gleser utility formula assuming normal performance distributions.

## Sturman (2000) Adjustments

**Important Note:** Our analysis applies a simplified 35% reduction factor, while @sturman2000implications actual findings were much more dramatic:

- **Our Adjusted Utility:** $`r formatC(round(sturman$total_utility), format = "d", big.mark = ",")`
- **Our Reduction:** $`r formatC(round(traditional$total_utility - sturman$total_utility), format = "d", big.mark = ",")` (35%)

**Sturman's Actual Findings (Original Latham & Whyte Case):**
- **Sturman's Adjusted Utility:** $2,228,170 (mean)
- **Sturman's Reduction:** $57,429,362 (**96.3% reduction**)
- **Median Effect Size:** 291% (across all adjustments)

Sturman's analysis revealed that utility estimates are highly contingent on organizational and environmental factors, with reductions ranging from 71% to over 96% depending on circumstances.

## Star Performer Approach
- **Star-Adjusted Utility:** $`r formatC(round(star_power$total_utility), format = "d", big.mark = ",")`
- **Improvement:** $`r formatC(round(star_power$total_utility - traditional$total_utility), format = "d", big.mark = ",")` (+175%)
- **SDy Multiplier:** 2.75× [@burke1986comparison]

Recognizes heavy-tailed performance distributions where star performers create exceptional value using SDy = 1.1 × mean salary [@joo2022hrm].

## Combined Approach
- **Combined Utility:** $`r formatC(round(combined$total_utility), format = "d", big.mark = ",")`
- **Net Improvement:** $`r formatC(round(combined$total_utility - traditional$total_utility), format = "d", big.mark = ",")` (+140%)

Balances Sturman's realism with star performer recognition, applying a moderate 20% reduction to star estimates.

# Methodological Comparison: Our Adjustments vs. Sturman's Findings

## Adjustment Severity Comparison

```{r adjustment-comparison}
# Create comparison with Sturman's actual findings
comparison_data <- data.frame(
  Scenario = c("Our Sturman Adjustment", "Sturman's Actual Mean", "Sturman's Actual Median"),
  Utility = c(formatC(round(sturman$total_utility), format = "d", big.mark = ","),
              "$2,228,170", "$1,738,861"),
  Reduction = c("35%", "96.3%", "97.1%"),
  Method = c("Simplified factor", "Complex Monte Carlo", "Complex Monte Carlo")
)

kable(comparison_data,
      col.names = c("Scenario", "Total Utility ($)", "Reduction (%)", "Method"),
      caption = "Comparison: Our Adjustments vs. Sturman's Actual Findings",
      align = c("l", "r", "r", "l")) %>%
  kable_styling(font_size = 10, latex_options = c("striped", "hold_position"))
```

## Key Differences

1. **Complexity:** Sturman used five specific adjustments (economic variables, multiple devices, top-down hiring, probationary period, employee flows) while we used a simplified reduction factor

2. **Magnitude:** Sturman's reductions were far more dramatic (>90%) than our conservative 35%

3. **Variability:** Sturman showed high variability across scenarios; our approach uses a fixed factor

4. **Realism:** Sturman's approach may be more realistic but requires extensive organizational data

# Strategic Implications

## Investment Justification

Even under conservative assumptions, the selection program generates substantial value:

- **Assessment Costs:** $`r formatC(round(lw_params$n_applicants * lw_params$cost_per_applicant), format = "d", big.mark = ",")`
- **Conservative ROI (Our Sturman):** `r round(sturman$total_utility / (lw_params$n_applicants * lw_params$cost_per_applicant))`× return
- **Realistic ROI (Combined):** `r round(combined$total_utility / (lw_params$n_applicants * lw_params$cost_per_applicant))`× return
- **Sturman's Actual ROI:** `r round(2228170 / (lw_params$n_applicants * lw_params$cost_per_applicant))`× return

## Research Evolution Timeline

```{r evolution-table}
evolution_data <- data.frame(
  Period = c("1940s-1980s", "2000s", "2020s"),
  Development = c("Classical utility analysis", "Sturman's operational realism", "Star performer recognition"),
  Key_Finding = c("Positive utility estimates", "Highly variable, often reduced", "Exceptional value of top talent"),
  Representative_Study = c("Brogden (1949)", "Sturman (2000)", "Joo et al. (2022)")
)

kable(evolution_data,
      col.names = c("Period", "Development", "Key Finding", "Representative Study"),
      caption = "Evolution of Utility Analysis Methodology",
      align = c("c", "l", "l", "l")) %>%
  kable_styling(font_size = 9, latex_options = c("striped", "scale_down"))
```

# Conclusions

This analysis demonstrates the evolution of utility analysis methodology and reveals important insights:

## Key Findings

1. **Traditional utility analysis** provides a useful baseline but may substantially overestimate benefits [@sturman2000implications]

2. **Sturman's adjustments** revealed utility estimates are highly contingent on organizational factors, with actual reductions often exceeding 90%

3. **Our simplified 35% reduction** represents a conservative approximation of Sturman's complex, scenario-dependent adjustments

4. **Star performer recognition** [@joo2022hrm] reveals exceptional value when organizations can identify top talent

5. **Combined approaches** offer balanced assessments recognizing both operational constraints and star performer potential

## Methodological Implications

- **Sturman's contribution:** Demonstrated that utility analysis requires careful consideration of organizational context
- **Star performer contribution:** Recognized that performance distributions are often heavy-tailed
- **Future research:** Need for integration of both realistic constraints and star performer recognition

## Practical Recommendations

Even under Sturman's most conservative scenarios, high-validity selection procedures remain valuable investments. The key is designing selection systems capable of identifying potential star performers while maintaining realistic expectations about organizational constraints.

# References 