---
title: "Reproduction of Seijts, Espinoza, & Carswell (2020): Utility Analysis of Character Assessment in Employee Placement"
author: "Utility Analysis Research Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Load required libraries
library(tidyverse)
library(psych)
library(corrplot)
library(ggplot2)
library(knitr)
library(kableExtra)
library(DT)

# Load results
load("seijts_2020_results.RData")

# Load synthetic data
synthetic_data <- read.csv("seijts_2020_synthetic_data.csv")
```

# Executive Summary

This report documents the reproduction of Seijts, Espinoza, and Carswell's (2020) utility analysis of character assessment in employee placement. The original study examined the economic value of using the Leader Character Insight Assessment (LCIA) for selecting senior managers, finding substantial returns on investment ranging from 15,637% to 23,505%.

## Key Findings

- **Original Study**: Reported utility of CAD $564,128 (15 years) and CAD $375,285 (10 years)
- **Our Reproduction**: Calculated utility of CAD $226,221 (15 years) and CAD $150,006 (10 years)
- **Discrepancy**: Our calculations are approximately 60% lower than reported values
- **Economic Adjustments**: Generated 81 different adjustment scenarios with utilities ranging from CAD $35,086 to CAD $2,173,042

## Methodological Insights

The discrepancy suggests potential differences in:
1. Zxs calculation methodology
2. Cost structure assumptions
3. Economic adjustment parameters
4. Incremental validity calculations

# Introduction

## Study Background

Seijts et al. (2020) conducted a utility analysis to estimate the economic value of using character assessment in employee placement decisions. The study used the Leader Character Insight Assessment (LCIA) to evaluate 111 senior managers in a Canadian manufacturing organization, finding a correlation of 0.30 between character and performance.

## Research Question

What is the potential dollar impact associated with the placement of managers based on the assessment of leader character using the LCIA?

## Original Findings

- **Correlation**: 0.30 (p < 0.01) between leader character and performance
- **Basic Utility**: CAD $564,128 over 15 years (CAD $37,609 yearly)
- **Conservative Utility**: CAD $375,285 over 10 years (CAD $37,529 yearly)
- **ROI**: 23,505% (15 years) to 15,637% (10 years)

# Methodology

## Data Generation

We created synthetic data matching the reported correlation matrix and descriptive statistics from Table 1 of the original study.

```{r data-validation}
# Display summary statistics comparison
summary_stats <- read.csv("seijts_2020_summary_stats.csv")
kable(summary_stats, digits = 3, 
      caption = "Summary Statistics: Reported vs Generated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Utility Analysis Framework

### Basic Brogden-Cronbach-Gleser Model

The basic utility formula used was:

$$\Delta U = N \times T \times r \times Z_{xs} \times SD_y - C$$

Where:
- $N$ = Number of candidates selected per year (1)
- $T$ = Average tenure (10 or 15 years)
- $r$ = Validity coefficient (0.30)
- $Z_{xs}$ = Average standardized predictor score of selected candidates
- $SD_y$ = Standard deviation of job performance in dollars (CAD $115,500)
- $C$ = Cost per candidate (CAD $800)

### Economic Adjustments

We implemented the adjusted utility formula from Seijts et al. (2020):

$$\Delta U = N \times T_{mod} \times (r_{old} - r_{new}) \times Z_{xs} \times [SD_y \times (1 + VC) \times (1 - TAX)] - \{N \times [C \times (1 - TAX)] / SR\}$$

Where additional parameters include:
- $T_{mod}$ = Modified tenure with discounting
- $VC$ = Variable costs rate
- $TAX$ = Taxation rate
- $r_{old} - r_{new}$ = Incremental validity

# Results

## Basic Utility Analysis

```{r basic-utility}
# Display basic utility results
basic_results <- data.frame(
  Tenure = c("15 years", "10 years"),
  Reported_Utility = c(seijts_2020_results$comparison$reported_15, 
                      seijts_2020_results$comparison$reported_10),
  Calculated_Utility = c(seijts_2020_results$basic_utility$utility_15,
                        seijts_2020_results$basic_utility$utility_10),
  Reported_Yearly = c(seijts_2020_results$comparison$reported_yearly_15,
                     seijts_2020_results$comparison$reported_yearly_10),
  Calculated_Yearly = c(seijts_2020_results$basic_utility$yearly_15,
                       seijts_2020_results$basic_utility$yearly_10),
  ROI_Reported = c(23505, 15637),
  ROI_Calculated = c(seijts_2020_results$basic_utility$roi_15,
                    seijts_2020_results$basic_utility$roi_10)
)

kable(basic_results, digits = 0, 
      caption = "Basic Utility Analysis: Reported vs Calculated Values") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Economic Adjustments Analysis

We generated 81 different scenarios by combining various adjustment parameters from Sturman (2000):

```{r adjustment-summary}
# Summary of economic adjustments
adjustment_summary <- data.frame(
  Tenure = c("15 years", "10 years"),
  Mean_Utility = c(mean(seijts_2020_results$results_15$utility),
                  mean(seijts_2020_results$results_10$utility)),
  Median_Utility = c(median(seijts_2020_results$results_15$utility),
                    median(seijts_2020_results$results_10$utility)),
  Min_Utility = c(min(seijts_2020_results$results_15$utility),
                 min(seijts_2020_results$results_10$utility)),
  Max_Utility = c(max(seijts_2020_results$results_15$utility),
                 max(seijts_2020_results$results_10$utility))
)

kable(adjustment_summary, digits = 0,
      caption = "Summary of Economic Adjustments (81 scenarios)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Distribution of Adjusted Utilities

```{r utility-distribution, fig.cap="Distribution of Utility Estimates Across 81 Adjustment Scenarios"}
# Create histogram of utility distributions
par(mfrow = c(1, 2))

hist(seijts_2020_results$results_15$utility / 1000, 
     main = "15-Year Tenure Utility Distribution",
     xlab = "Utility (CAD $ thousands)", 
     ylab = "Frequency",
     col = "lightblue", border = "black")

hist(seijts_2020_results$results_10$utility / 1000,
     main = "10-Year Tenure Utility Distribution", 
     xlab = "Utility (CAD $ thousands)",
     ylab = "Frequency",
     col = "lightgreen", border = "black")
```

## Parameter Sensitivity Analysis

```{r sensitivity-analysis, fig.cap="Sensitivity of Utility Estimates to Adjustment Parameters"}
# Create boxplots showing sensitivity to different parameters
library(ggplot2)

# Prepare data for plotting
plot_data_15 <- seijts_2020_results$results_15 %>%
  gather(key = "Parameter", value = "Value", VC, TAX, DISCOUNT, INCREMENTAL) %>%
  mutate(Parameter = factor(Parameter, 
                           levels = c("VC", "TAX", "DISCOUNT", "INCREMENTAL"),
                           labels = c("Variable Costs", "Taxation Rate", 
                                    "Discount Rate", "Incremental Validity")))

ggplot(plot_data_15, aes(x = Parameter, y = utility/1000, fill = Parameter)) +
  geom_boxplot() +
  labs(title = "Sensitivity of 15-Year Utility to Adjustment Parameters",
       y = "Utility (CAD $ thousands)",
       x = "Parameter") +
  theme_bw() +
  theme(legend.position = "none",
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white")) +
  coord_flip()
```

# Discussion

## Key Discrepancies

Our reproduction revealed significant differences from the reported values:

1. **Basic Utility**: Our calculations are approximately 60% lower than reported
2. **Zxs Calculation**: We used qnorm(1-selection_ratio) = 0.44, but the paper may have used different methodology
3. **Cost Structure**: Potential differences in how costs were calculated or allocated

## Potential Explanations

### Zxs Calculation Differences

The paper mentions using "Naylor and Shine's (1965) tables for determining the increase in mean criterion score (1.09)" but our calculation yielded 0.44. This suggests:

1. Different selection ratio interpretation
2. Use of different tables or methodology
3. Potential rounding or approximation differences

### Economic Adjustments

The paper reports using "highly conservative adjustments" that still yield positive returns. Our analysis shows:

1. Wide range of possible utilities (CAD $35K to CAD $2.2M)
2. Most scenarios still yield positive returns
3. Conservative scenarios align more closely with reported values

## Methodological Insights

### Strengths of the Original Study

1. **Multi-source data**: Peer ratings for character, supervisor ratings for performance
2. **Comprehensive adjustments**: Multiple economic factors considered
3. **Conservative approach**: Multiple adjustment scenarios tested

### Limitations and Considerations

1. **Single organization**: Results may not generalize
2. **Concurrent validity**: Causality cannot be established
3. **Parameter assumptions**: Many economic parameters estimated rather than measured

## Implications for Practice

Despite the discrepancies, our reproduction confirms:

1. **Positive returns**: Character assessment shows economic value
2. **Parameter sensitivity**: Results highly dependent on assumptions
3. **Conservative estimates**: Even with adjustments, positive ROI likely

# Conclusion

## Reproduction Success

We successfully reproduced the core methodology of Seijts et al. (2020) and generated synthetic data matching their reported statistics. While our basic utility calculations differ from reported values, the economic adjustments analysis demonstrates that character assessment can provide substantial economic value.

## Key Contributions

1. **Methodological transparency**: Detailed documentation of utility calculations
2. **Parameter sensitivity**: Comprehensive analysis of adjustment effects
3. **Validation framework**: Systematic approach to reproducing utility analyses

## Future Directions

1. **Zxs methodology**: Investigate different approaches to calculating standardized predictor scores
2. **Cost structure**: Examine alternative cost allocation methods
3. **Cross-validation**: Test methodology with additional datasets

## Recommendations

1. **Conservative estimates**: Use lower-bound utility estimates for planning
2. **Sensitivity analysis**: Always test multiple parameter combinations
3. **Transparency**: Document all assumptions and calculation methods

---

**Note**: This reproduction study was conducted for educational and research purposes. All calculations and interpretations should be verified independently before use in practice.

# References

Seijts, G., Espinoza, J. A., & Carswell, J. (2020). Utility analysis of character assessment in employee placement. *Leadership & Organization Development Journal*, 41(5), 703-720.

Sturman, M. C. (2000). Implications of utility analysis adjustments for estimates of human resource intervention value. *Journal of Management*, 26(2), 281-299.

Boudreau, J. W. (1983a). Economic considerations in estimating the utility of human resource productivity improvements. *Personnel Psychology*, 36(3), 551-557. 